---
title: "Barplot in R Markdown"
author: "Dr. Suborna Ahmed"
output:
  word_document: default
  html_document:
    df_print: paged
---


# Preparing directory and data setup
In this lecture, we will be creating side-by-side graphs for two different species using the vri.csv file. For doing this use the <code>subset()</code> and <code>par()</code> functions such as were explained in previous lectures.

```{r}
# VRI data

vri_data<-read.csv("VRI_data.csv",header = T)

attach(vri_data)

# separating data by species
d1<- subset(vri_data, (SPECIES_CD_1 =='CW'))
d2<- subset(vri_data, (SPECIES_CD_1 =='PLC'))
```

# Add one legend for multiple graphs
We can add a legend for multiple graphs by using the <code>legend()</code> function. In this case, we have to use the last graph created as a reference to add the coordinates of the legend. Make sure that you give appropriate margin space in the graph for adding the legend using the <code>mai=</code> option in the <code>par()</code> function. Use <code> xpd=NA</code> to specify that we want the legend outside of the graph area.

```{r}


ylabel="Projected Height (m)"
xlabel="Projected Age (years)"
ylm=c(0,60)
xlm=c(0,470)

cxlb=1.1 # axis level - fontsize
cxaxis=1.1 # axis fontsize
maincx=2.5 # title font

row1=3 # age column number
row2=4 # ht column number
ldcex=1 # size ofpoints
ld=1
maxd1<- max(d2[,row2])
m1<-subset(d2, PROJ_HEIGHT_1==maxd1)
maxd2<- max(d1[,row2])
m2<-subset(d1, PROJ_HEIGHT_1==maxd2)
#names(vri_data)

# creating a permanent image file
png("One_legend.png", width=5, height=5, units = 'in', res=1200)
par(mfrow=c(1,2),mai=c(3,0.9,0.2,0.3), cex=1.0, xpd=NA) # mai is the margin

# graph 1
plot(d1[,row1],d1[,row2], type = "p", pch = 20, col = "blue",  ylim=ylm, xlim=xlm, cex=ldcex, cex.lab=cxlb, cex.axis=cxaxis,  xlab=xlabel, ylab=ylabel)
points(m2[,row1],m2[,row2], type = "p", pch =11, col = "red", lwd=ld)

# graph 2
plot(d2[,row1],d2[,row2], type = "p", pch = 20, col = "black",  ylim=ylm, xlim=xlm, cex=ldcex, cex.lab=cxlb, cex.axis=cxaxis,  xlab=xlabel, ylab=ylabel)
points(m1[,row1],m1[,row2], type = "p", pch =11, col = "red", lwd=ld)

legend(-700, -35, ncol=1 ,c("Projected height: Species is 'CW'", "Projected height: Species is 'PLC'", "Maximum Projected Height"), col = c("blue", "black", "red"), pch = c(20, 20, 11), cex=1, pt.cex=ld)

graphics.off()

```

# Creating barplots
For this section we will create a bar plot using the trees.txt file. Use the <code> read.table()</code>  function for importing this file into R. Then:

1.	Determine the type of variables that you have (categorical or numerical). In this case we want to know what type of variable is “CC”. Use the <code> class()</code>  function for checking on this.
2.	 Use <code> table()</code>  to create a vector with the counts of the number of items that you have in the CC column.
3.	Use <code> barplot()</code>  function to create a bar plot. Use the vector that you create with the counts as a file and modify graph limits, colours, and labels if desired. 

```{r}
# .txt data file
trees <- read.table("trees.txt", header = TRUE)
# checking the data:
dim(trees)
names(trees)
# attached all variables
attach(trees)

# crown class variable and create a table
class(CC)
counts <- table(CC)
counts

#barplot 1
par(mfrow=c(1,1),mai=c(1,0.9,0.2,0.3), cex=1.0, oma=c(5,2,2,0), xpd=NA) # mai is the margin
barplot(counts)
graphics.off()

#barplot 2
graphics.off()
par(mfrow=c(1,1),mai=c(1,0.9,0.2,0.3), cex=1.0, oma=c(5,2,2,0), xpd=NA) # mai is the margin
barplot(counts, ylim = c(0, 20))

#barplot 3
graphics.off()
par(mfrow=c(1,1),mai=c(1,0.9,0.2,0.3), cex=1.0, oma=c(5,2,2,0), xpd=NA) # mai is the margin
barplot(counts, ylim = c(0, 20), ylab = "Frequency", xlab = "Crown Classes")

detach(trees)

```


# VRI data bar chart
We can also create bar plots by selecting a categorical variable and using the <code>  plot()</code>   function. In this case, R will automatically create a bar plot with the selected categorical variable (e.g. <code>plot (SPECIES_CD_1) </code>). To change the direction of the numbers in the y axis use <code>  las=1</code> into <code>barplot()</code>. To change the names in the x axis use <code>names.arg= c(“name_variable1”, “name_variable2”)</code>. This will change the names that you have for each column in the x axis. To modify the font size use <code>cex.names=</code>.

```{r}
attach(vri_data)
class(SPECIES_CD_1)
 
#barplot 1
#coverting the character variable as a factor to plot it

plot(factor(SPECIES_CD_1))

#barplot 2
leading_spec <- table(SPECIES_CD_1)
percent_spec <- (table(SPECIES_CD_1)*100)/2700
barplot(percent_spec)

#barplot 3
barplot(percent_spec, main="Leading Species of Stands", 	xlab="Species", ylab="%")

#barplot 4: rotate y axis values:

barplot(percent_spec, main="Leading Species of Stands", 	xlab="Species", ylab="%", las=1)

#barplot 5
# set the names of categories under x axis:
barplot(percent_spec, main="Leading Species of Stands", 	xlab="Species", ylab="%", las=1, names.arg=c("non-treed stands",  "Western Red Cedar", 	"Alder", "Western Hemlock", "Lodgepole pine",  "Sitka Spruce", "Yellow 	Cedar"))

# barplot 6: reduce the names to adjust the space
barplot(percent_spec, main="Leading Species of Stands", 	xlab="Species", ylab="%", las=1, names.arg=c("Non-treed",  "Western R. Cedar", 	"Alder", "Western Hemlock", "L. pine",  "Sitka Spruce", "Yellow 	Cedar"), ylim=c(0,60), cex.axis=1.5, cex.lab=1.5, cex.names=0.7)


```

